Page not found

### 2023-06-21 manuscript/appendices/A_fixed_points.qmd
so... many.... figures !
proofread appendix

We recall the equation for the low-rank network of neurons :

$$
\dot h_i(t) = -h_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^N F_{\mu i} G_{\mu j} \phi(h_j(t))
$$

@@

We note that in the main text, the distribution is a multivariate normal with unit covariance and zero mean, such that it factors out as $\rho(z_1, \cdots, z_p) = \prod_{\mu=1}^p \mathcal{N}(z_\mu)$. Furthermore, the functions are component-wise $f_\mu(z_1, \cdots, z_p) = z_\mu$ and $g_\mu(z_1, \cdots, z_p) = \tilde \phi(z_\mu) = \frac{\phi(z_\mu) - \avg{\phi(z_\mu)}}{\mathrm{Var}[\phi(z_\mu)]}$.

@@

$$
\avg{g_\mu(\vec{z})} = \avg{\tilde \phi(z_\mu)} = \avg{\frac{\phi(z_\mu) - \avg{\phi(z_\mu)}}{\mathrm{Var}[\phi(z_\mu)]}} = 0
$$

@@

Another set of fixed points is $h^\star_i = F_{\nu i}$ for all $\nu = 1, \cdots, p$. We refer to these as the "patterns" in the context of learning the network "remembers" (converges to) these vectors. We verify the fixed point similarly as before.

@@

As $N \to \infty$, the fixed points $h_i^\star$ become functions $h^\star(z_1, \cdots, z_p)$ on the $\Rp$ space :

@@

We now study the stability of the fixed points. In the original equation, we substitute $h_i(t) = h_i^\star + \delta_i(t)$, where $\delta_i(t)$ is a small perturbation.

@@

Before we study the spectrum of $K$, we cite the following result: if $\lambda$ is an eigenvalue of $A$, then $\lambda-1$ is an eigenvalue of $A-\mathrm{Id}$. This follows simply from the equation for the eigenvalues $\lambda_i$ and eigenvectors $\vec{v_i}$ of $A$. If $A \vec{v_i} = \lambda_i \vec{v_i}$, then $(A-\mathrm{Id})\vec{v_i}=\lambda_i \vec{v_i} - \vec{v_i} = (\lambda_i - 1)\vec{v_i}$. Therefore to study $K$ it suffices to study the eigenvalues of $J^\prime$.

@@

The term $\avg{g^\prime_\nu(z_\nu) f_\nu(z_\nu)}$ does not vanish in general because of the coupling between the components. The integral is too complicated to be analytically solved, but we provide a numerical estimation in our case ($\rho(z_1, \cdots, z_p)=\prod_{\mu=1}^p\mathcal{N}(z_\mu)$, $f_\mu(z_\mu)=z_\mu$, $g_\mu(z_\mu)=\tilde\phi(z_\mu)$) and with fixed points $h_i^\star=0$ and $h_i^\star=F_{\nu i}$.

@@

We now consider a Taylor expansion of the rate function $\phi(h)$ up to order 3, and cite results obtained from symbolic computation and known moments of the Gaussian distribution.

@@

What a mess ! We summarize the results from the Taylor expansion :

* When $\phi$ is linear, we see that all the eigenvalues are zero, therefore the first order stability is inconclusive. Numerically, we see that when $\phi$ is linear, the convergence to the fixed points is very slow.
* When $\phi$ is quadratic, the zero fixed point is stable, and the pattern fixed point is unstable, regardless of the sign of the coefficients.
* When $\phi$ is cubic, the sign of the eigenvalues depends primarily on $c_3$ (negative $c_1$ breaks the assumption that $\phi$ must be monotonically increasing). The intuition is that we require $c_1$ to be large enough (in other words, the growth of the firing rate when $h$ becomes positive) and $c_3 < 0$ such that $3 c_1 c_3 + 15 c_3^2 < 0$.
  The intuition behind this is that if the slope $c_1$ is stronger, then the bump of $\partial \phi$ becomes "more concentrated", which helps lower the eigenvalues associated with the pattern fixed points.
  The magnitude of the second order term $c_2$ only contributes to adding instability to the pattern fixed point, and stability to the zero fixed point, which leads to the conclusion it is best set to $c_2=0$ if we want the network to be stable at the pattern fixed points.

[TODO FIG : might be nice to have a phase plot of $c_1, c_3$ and the value of the eigenvalue to illustrate the previous illustration]

The study of the Taylor expansion therefore instructs us that stability is improved when $\phi$ grows rapidly around $h=0$, and that even powers (symmetries of $\phi$ around the y-axis) deteriorate the stability of the pattern fixed point. This helps us understand why the logistic function $\phi(h)=\frac{1}{1+\mathrm{e}^{-h}}=\tfrac 12 + \tfrac h4 - \tfrac{h^3}{48} + \bO(h^5)$ has good convergence properties : we have that $c_1$ is large enough, $c_3$ is negative, and $c_2$ is zero.

@@

In conclusion, the spectrum of $K$ is composed of $p$ eigenvalues $\lambda_\nu = \avg{g_\nu(z_\nu) f_\nu(z_\nu) \partial \phi(h^\star)}-1$, and $N-p$ eigenvalues $\lambda_\perp = -1$. For the network considered in this work, the interpretation of this is that components of the field aligned with the patterns $z_\nu$ converge to the fixed point, and the orthogonal components decay as $t \to \infty$. The fixed point $h_i^\star=0$ is therefore unstable, and the pattern fixed points $h_i^\star = z_{\nu i}$ are stable. We also note that since the normal distribution is centred around zero and symmetric, $h_i^\star = -z_{\nu i}$ are also stable fixed points.

>>>>

We recall the equation for the low-rank network of neurons:

$$
\dot h_i(t) = -h_i(t) + \frac 1N \sum_{\mu=1}^p \sum_{j=1}^N F_{\mu i} G_{\mu j} \phi(h_j(t)).
$$

@@

In the main text, the distribution is a multivariate normal with unit variance, no covariance and zero mean, such that it factors out as $\rho(z_1, \cdots, z_p) = \prod_{\mu=1}^p \mathcal{N}(z_\mu)$. Furthermore, the functions are component-wise $f_\mu(z_1, \cdots, z_p) = z_\mu$ and $g_\mu(z_1, \cdots, z_p) = \tilde \phi(z_\mu) = \frac{\phi(z_\mu) - \avg{\phi(z_\mu)}}{\mathrm{Var}[\phi(z_\mu)]}$.

@@

$$
\avg{g_\mu(\vec{z})} = \avg{\tilde \phi(z_\mu)} = \avg{\frac{\phi(z_\mu) - \avg{\phi(z_\mu)}}{\mathrm{Var}[\phi(z_\mu)]}} = 0.
$$

@@

Another set of fixed points is $h^\star_i = F_{\nu i}$ for all $\nu = 1, \cdots, p$. We refer to these as the "patterns", since in the context of learning the network "remembers" (converges to) these vectors. We verify the fixed point similarly as before.

@@

As $N \to \infty$, the fixed points $h_i^\star$ become functions $h^\star(z_1, \cdots, z_p)$ on the $\Rp$ space:

@@

We now study the stability of the fixed points. In the evolution equation, we substitute $h_i(t) = h_i^\star + \delta_i(t)$, where $\delta_i(t)$ is a small perturbation.

@@

Before we study the spectrum of $K$, we write the following result: if $\lambda$ is an eigenvalue of $A$, then $\lambda-1$ is an eigenvalue of $A-\mathrm{Id}$. This follows simply from the equation for the eigenvalues $\lambda_i$ and eigenvectors $\vec{v_i}$ of $A$. If $A \vec{v_i} = \lambda_i \vec{v_i}$, then $(A-\mathrm{Id})\vec{v_i}=\lambda_i \vec{v_i} - \vec{v_i} = (\lambda_i - 1)\vec{v_i}$. Therefore, to study $K$, it suffices to study the eigenvalues of $J^\prime$.

@@

The term $\avg{g^\prime_\nu(z_\nu) f_\nu(z_\nu)}$ does not vanish in general because of the coupling between the components. The integral is too complicated to be analytically solved, but we provide a numerical estimation in our case ($\rho(z_1, \cdots, z_p)=\prod_{\mu=1}^p\mathcal{N}(z_\mu)$, $f_\mu(z_\mu)=z_\mu$, $g_\mu(z_\mu)=\tilde\phi(z_\mu)$) for fixed points $h_i^\star=0$ and $h_i^\star=F_{\nu i}$.

@@

We now consider a Taylor expansion of the activation function $\phi(h)$ up to order 3, and cite results obtained from symbolic computation and known moments of the Gaussian distribution.

@@

What a mess! We summarize the results from the Taylor expansion:

* When $\phi$ is linear, we see that all the eigenvalues are zero, therefore the first order stability is inconclusive. Simulations show when $\phi$ is linear, the convergence to the fixed points is very slow (the spectrum is negative, but the eigenvalues $\lambda_\nu$ are weakly negative, and tend to zero when $N \to \infty$).
* When $\phi$ is quadratic, the zero fixed point is stable, and the pattern fixed point is unstable, regardless of the sign of the coefficients.
* When $\phi$ is cubic, the sign of the eigenvalues depends primarily on $c_3$ (negative $c_1$ breaks the assumption that $\phi$ must be monotonically increasing). The intuition is that we require $c_1$ to be large enough (in other words, the growth of the firing rate when $h$ becomes positive) and $c_3 < 0$ such that $3 c_1 c_3 + 15 c_3^2 < 0$.
  The intuition behind this is that if the slope $c_1$ is stronger, then the bump of $\partial \phi$ becomes "more concentrated", which helps lower the eigenvalues associated with the pattern fixed points.
  The magnitude of the second order term $c_2$ only contributes to adding instability to the pattern fixed point, and stability to the zero fixed point, which leads to the conclusion it is best set to $c_2=0$ if we want the network to be stable at the pattern fixed points.

<!-- [TODO FIG: might be nice to have a phase plot of $c_1, c_3$ and the value of the eigenvalue to illustrate the previous illustration] -->

The study of the Taylor expansion therefore instructs us that stability is improved when $\phi$ grows rapidly around $h=0$, and that even powers (symmetries of $\phi$ around the y-axis) deteriorate the stability of the pattern fixed point. This helps us understand why the logistic function $\phi(h)=\frac{1}{1+\mathrm{e}^{-h}}=\tfrac 12 + \tfrac h4 - \tfrac{h^3}{48} + \bO(h^5)$ has good convergence properties: we have that $c_1$ is large enough, $c_3$ is negative, and $c_2$ is zero.

@@

In conclusion, the spectrum of $K$ is composed of $p$ eigenvalues $\lambda_\nu = \avg{g_\nu(z_\nu) f_\nu(z_\nu) \partial \phi(h^\star)}-1$, and $N-p$ eigenvalues $\lambda_\perp = -1$. For the network considered in this work, the interpretation is that components of the field aligned with the patterns $z_\nu$ converge to the fixed point, and the orthogonal components decay as $t \to \infty$. The fixed point $h_i^\star=0$ is therefore unstable, and the pattern fixed points $h_i^\star = z_{\nu i}$ are stable. We finally note that since the normal distribution is centred around zero and symmetric, $h_i^\star = -z_{\nu i}$ are also stable fixed points.

### 2023-06-21 manuscript/appendices/B_mean_patterns.qmd
proofread appendix

For now, we do not use the fact that in our setting, $J_{ij}$ is low-rank.
Applying the finite-$n$ mapping $S^n$,[^details] the neurons $\A = \{i_1, \cdots, i_{|\A |}\}$ will end up in the same 1D bin at position $\alpha$. There are 1D $4^n$ bins, each of length $4^{-n}$.

@@

As explained in the main text, we average the neurons together in a bin, and we write the corresponding potential of the segment population at location $\alpha$:

@@

We now make the mean-field approximation by dropping $\sum_{i \in \A}$, that is, that all the neurons inside of each bin are the same (or at least, sufficiently similar so that the mean-field is not a big mistake). Doing so, we recover @eq-evolution-network, which concludes the proof.

@@

@eq-Jab-mean-lr also nicely complements the intuition that the more neurons are in each bin, the more "important" this bin should be. This is expressed through the weights $\frac{|\B|}{\sum_{\beta^\prime} |\B^\prime|}$.

@@

For the sake of illustration, let us compare the connectivity kernels obtained using different methods in @fig-kernels.

* @fig-kernels-1 is the method described in this appendix, based on sampling $N$ neurons in the PDF space $\R^2$, applying the mapping on the obtained positions $\vec{z_i}$, the applying the binning (downsampling). We see, through the empty rows in the connectivity, that due to finite-$N$ effect, locations where $\rho$ is small are not represented, and, as a consequence, no neurons are present in the corresponding 1D bin. When this happens, we set these rows to zero.
* To obtain @fig-kernels-2, we make the approximation that the support of $\rho(z_1, z_2)$ is compact in $[-4,4]\times[-4,4]$, which of course is not true, but in the case of a Gaussian distribution this is a reasonable approximation, because $\CDF(4) \approx 0.999968$. Defining $\tilde \rho = \rho \circ S^{-1}$, the grid is numerically renormalized so that despite the compact support approximation, the probabilities sum up to one, and the connectivity matrix is modified so that the columns are weighed by $\tilde \rho(\beta)$:
  $$
  \tilde J_{\alpha\beta} = \tilde \rho(\beta) \sum_{\mu=1}^p \tilde F_{\mu \alpha} \tilde G_{\mu \beta}
  $$
  We see that compared to @fig-kernels-1, using the grid method allows us to "sample" the locations where the density is small, and this results in less "whited-out" rows. The vertical bands of white correspond to the reweighing of the columns with $\tilde \rho(\beta)$.
* Finally, @fig-kernels-3 shows the kernel obtained from applying the grid method in the CDF $[0,1]^2$ described in @sec-mapping-cdf. We see that the connectivity seems much more "uniform", because each point of the grid has "equal weight" in the CDF space.

>>>>

For now, we do not use the fact that in our setting, $J_{ij}$ is low-rank.
Applying the finite-$n$ mapping $S^n$,[^details] the neurons $\A = \{i_1, \cdots, i_{|\A |}\}$ will end up in the same 1D bin at position $\alpha$. We stress that $|\A| \neq 2^n$ in general, because we sample the distribution, instead of applying a grid. There are, however, still $4^n$ 1D bins, each of length $4^{-n}$.

@@

As explained in the main text, we average the segment together in a bin, and we write the corresponding potential of the segment population at location $\alpha$:

@@

We now make the mean-field approximation by dropping $\sum_{i \in \A}$, that is, that all the neurons inside of each bin are the same (or at least, sufficiently similar so that the mean-field approximation is not a big mistake, in the sense that there is a lot of variation inside each bin, see @sec-locality). Doing so, we recover @eq-evolution-network, which concludes the proof.

@@

@eq-Jab-mean-lr also nicely complements the intuition that the more neurons there are in each bin, the more "important" this bin should be. This is expressed through the weights $\frac{|\B|}{\sum_{\beta^\prime} |\B^\prime|}$.

@@

For the sake of illustration, let us compare the connectivity kernels obtained using different methods in @fig-kernels. These three approximations are equivalent, since they yield the same dynamics in the $N \to \infty$ limit.

* @fig-kernels-1 is the method described in this appendix, based on sampling $N$ neurons in the PDF space $\R^2$, applying the mapping on the obtained positions $\vec{z_i}$, and finally applying the binning (this is coarse-graining). We see, through the empty rows in the connectivity, that due to finite-$N$ effects, locations where $\rho$ is small are not represented, and, as a consequence, no neurons are present in the corresponding 1D bin. When this happens, we set these rows to zero.
* To obtain @fig-kernels-2, we make the approximation that the support of $\rho(z_1, z_2)$ is compact in $[-4,4]\times[-4,4]$, which of course is not true, but in the case of a Gaussian distribution this is a reasonable approximation, because $\CDF(4) \approx 0.999968$. Defining $\tilde \rho = \rho \circ S^{-1}$, the grid is numerically renormalized so that despite the compact support approximation, the probabilities sum up to one, and the connectivity matrix is modified so that the columns are weighed by $\tilde \rho(\beta)$:
  $$
  \tilde J_{\alpha\beta} = \tilde \rho(\beta) \sum_{\mu=1}^p \tilde F_{\mu \alpha} \tilde G_{\mu \beta}
  $$
  We see that compared to @fig-kernels-1, using the grid method allows us to "sample" the locations where the density is small, and this results in less "whited-out" *rows*. Still, vertical bands of very fade colour correspond to the reweighing of the *columns* where $\tilde \rho(\beta)$ is small.
* Finally, @fig-kernels-3 shows the kernel obtained from applying the grid method in the CDF space $[0,1]^2$, as described in @sec-mapping-cdf. We see that the connectivity seems much more "uniform", because each point of the grid has "equal weight" in the CDF space.
